---
title: "Midterm Project Report"
author: "Maddy Rilling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
options(warn = -1)
```

## How has the location of shots in the NBA changed over the years?
```{r, echo=FALSE, message=FALSE}
## Load the necessary libraries
library(devtools)
library(ggplot2)
library(nbastatR)

## Install the package nbastatR from GitHub
devtools::install_github("abresler/nbastatR")

## Make buffer size larger using the vroom package to allow the large data file to load correctly 
Sys.setenv("VROOM_CONNECTION_SIZE" = 262144)

## Create variable years to get the wanted seasons for graphic
years <- c(2000, 2010, 2020)

## Initialize an empty dataframe to store the combined shot data for the three seasons
combined_shot_data <- data.frame()

## Loop through each season
for (year in years) {
  ## Retrieve shot data for Milwaukee Bucks for specificed seasons
  shot_data <- teams_shots(teams = "Milwaukee Bucks", seasons = year)
  ## Use the rbind function to combine shot data for each season into one dataframe
  combined_shot_data <- rbind(combined_shot_data, shot_data)
}

## Create the graphical representation of the shot data and store in object called shot_chart
## Use locationX/10 because need to convert the data from feet to units
## Use locationY/10 because need to convert the data from feet to units and shift the data down 42 units
## Color of data points based on if the shot taken was made or missed
shot_chart <- ggplot(combined_shot_data, aes(x = locationX/10, y = locationY/10-42, color = typeEvent)) +
  ## Alpha for opacity of data points and size to reflect size of data points
  geom_point(alpha = 0.4, size = 1) + 
  ## Color of data points plotted with labels
  scale_color_manual(values = c("#00471B", "#D2B48C"), name = "Shot Made", labels = c("Made","Missed")) +
  ## Labels for the graph
  labs(title = "Milwaukee Bucks Shot Chart",
       x = "Baseline",
       y = "Sideline") +
  ## Faceting to plot each season's short chart next to each other
  facet_wrap(~ yearSeason, nrow = 1) +
  ## Change theme
  theme_minimal() +
  ## Fill theme with white panel background
  theme(panel.background = element_rect(fill = "white"))

## After plotting the shot chart, add the full basketball court on top of the shot chart points
## I used the code from GitHub for the full court plot, but modified it to use annotate so that it would plot on top of the shot chart
## Full court code from Github user edkupfer (https://gist.github.com/edkupfer/6354964).
## This way, the full court is able to be viewed even with the shot chart data on top of it.
shot_chart +
  annotate("path", x = -25, y = c(-47, 47, 47, -47, -47), color = "black") +
  annotate("path", x = c(-25, 25), y = c(0, 0), color = "black") +
  annotate("path", x = c(-6000:(-1)/1000, 1:6000/1000), y = c(sqrt(6^2 - c(-6000:(-1)/1000,    1:6000/1000)^2)), color = "black") +
  annotate("path", x = c(-6000:(-1)/1000, 1:6000/1000), y = -c(sqrt(6^2 - c(-6000:(-1)/1000, 1:6000/1000)^2)), color = "black") +
  annotate("path", x = c(-6000:(-1)/1000, 1:6000/1000), y = c(28 - sqrt(6^2 - c(-6000:(-1)/1000, 1:6000/1000)^2)), color = "black") +
  annotate("path", x = c(-6000:(-1)/1000, 1:6000/1000), y = -c(28 - sqrt(6^2 - c(-6000:(-1)/1000, 1:6000/1000)^2)), color = "black") +
  annotate("path", x = c(-6000:(-1)/1000, 1:6000/1000), y = c(28 + sqrt(6^2 - c(-6000:(-1)/1000, 1:6000/1000)^2)), color = "black", linetype = "dashed") +
  annotate("path", x = c(-6000:(-1)/1000, 1:6000/1000), y = -c(28 + sqrt(6^2 - c(-6000:(-1)/1000, 1:6000/1000)^2)), color = "black", linetype = "dashed") +
  annotate("path", x = c(-8, -8, 8, 8, -8), y = c(47, 28, 28, 47, 47), color = "black") +
  annotate("path", x = -c(-8, -8, 8, 8, -8), y = -c(47, 28, 28, 47, 47), color = "black") +
  annotate("path", x = c(-6, -6, 6, 6, -6), y = c(47, 28, 28, 47, 47), color = "black") +
  annotate("path", x = c(-6, -6, 6, 6, -6), y = -c(47, 28, 28, 47, 47), color = "black") +
  annotate("path", x = c(-4000:(-1)/1000, 1:4000/1000), y = c(41.25 - sqrt(4^2 - c(-4000:(-1)/1000, 1:4000/1000)^2)), color = "black") +
  annotate("path", x = c(-4000:(-1)/1000, 1:4000/1000), y = -c(41.25 - sqrt(4^2 - c(-4000:(-1)/1000, 1:4000/1000)^2)), color = "black") +
  annotate("path", x = c(-750:(-1)/1000, 1:750/1000, 750:1/1000, -1:-750/1000), y = c(c(41.75 + sqrt(0.75^2 - c(-750:(-1)/1000, 1:750/1000)^2)), c(41.75 - sqrt(0.75^2 - c(750:1/1000, -1:-750/1000)^2))), color = "black") +
  annotate("path", x = c(-750:(-1)/1000, 1:750/1000, 750:1/1000, -1:-750/1000), y = -c(c(41.75 + sqrt(0.75^2 - c(-750:(-1)/1000, 1:750/1000)^2)), c(41.75 - sqrt(0.75^2 - c(750:1/1000, -1:-750/1000)^2))), color = "black") +
  annotate("path", x = c(-3, 3), y = c(43, 43), color = "black", lineend = "butt") +
  annotate("path", x = c(-3, 3), y = -c(43, 43), color = "black", lineend = "butt") +
  annotate("path", x = c(-22, -22, -22000:(-1)/1000, 1:22000/1000, 22, 22), y = c(47, 47 - 169/12, 41.75 - sqrt(23.75^2 - c(-22000:(-1)/1000, 1:22000/1000)^2), 47 - 169/12, 47), color = "black") +
  annotate("path", x = c(-22, -22, -22000:(-1)/1000, 1:22000/1000, 22, 22), y = -c(47, 47 - 169/12, 41.75 - sqrt(23.75^2 - c(-22000:(-1)/1000, 1:22000/1000)^2), 47 - 169/12, 47), color = "black") +
  annotate("path", x = 25, y = c(-47, 47, 47, -47, -47), color = "black") +
  annotate("path", x = c(-25, 25), y = c(-47, -47), color = "black") +
  annotate("path", x = c(-25, 25), y = c(47, 47), color = "black") +
  coord_fixed()

## Finding the optimal distance for the 4 point arc
## Calculate the distance of each shot from the basket using the Pythagorean theorem
combined_shot_data$distance_ft <- sqrt((combined_shot_data$locationX * 10)^2 + (combined_shot_data$locationY * 10)^2)

## Aggregate by the isShotMade variable
## combined_shot_data$isShotMade is used to tell whether each shot was a make or miss
## list(combined_shot_data$distance_ft) specifies which grouping variable is being used
## FUN = mean computes the mean of isShotMade for each distance range which gives us shooting efficiency
shooting_efficiency <- aggregate(combined_shot_data$isShotMade, 
                                 list(combined_shot_data$distance_ft), 
                                 FUN = mean)
## Calculate the expected value for each distance range
## Two-point shot range
shooting_efficiency$expected_value_2pt <- shooting_efficiency$x * 2  
## Three-point shot range
shooting_efficiency$expected_value_3pt <- shooting_efficiency$x * 3  

## Combine expected values for two-point and three-point shots
## Use maximum expected value for each distance range
shooting_efficiency$expected_value <- pmax(shooting_efficiency$expected_value_2pt, shooting_efficiency$expected_value_3pt)

## Find the distance range that gives the highest expected value
optimal_distance <- shooting_efficiency[which.max(shooting_efficiency$expected_value), "Group.1"]
```

### Analysis of Shot Charts

-From the graphics above, we can see that throughout the years more shots are taken outside of the three point line than before. 

-From the charts above, we can also see that the shots taken gradually get farther away from the basket overtime.

-A low percentage of shots taken past half court are made. 

-Many of the shots attempted within a small range of the basketball hoop are made shots.

-Based on where there is a high percentages of shots taken from can lend insight as to how opponents defense was set up. 

### More In-Depth Report

-Looking at the graphics above, there is a clear trend that more shots are taken outside of the three point line than seasons past. Many of the shots taken during the 2000 season, are inside the three point arc. As we move to the 2010 season, we can see that the shots taken are beginning to move closer to the three point arc. Lastly, the 2020 season shows us that there are many more shots taken outside of the three point arc, than the past seasons of 2000 and 2010. We can see that the area inside of the three point arc has less shots taken then the past seasons. This demonstrates the trend that as the seasons go on, more shots are being taken outside of the three point arc than the previous season.

-Another trend to be noted from the graphics above is that as the seasons go on, the shots begin to get further away from the basketball hoop. If we begin by looking at the shot chart from the 2000 season, we can see that many of the shots are taken in the basketball lane and outside of the lane as well. Many of the shots during this season stay about 2 or 3 feet away from the three point arc, which is shown in the graphic above. Moving to the 2010 season, we can see that the shots have moved closer to the three point arc than the 2000 season. During the 2010 season there is an area about 1 foot in front of the three point arc that not many shots are attempted at. Lastly, looking at the 2020 season shot chart, we see that many of the shots that were close to the three point arc but not behind it have moved completely behind the three point arc. The number of shots taken inside the three point arc have decreased from the season previous to these. This is because more shots are being taken outside of the three point arc than in previous seasons,leading to the decline of shots taken inside the three point arc. 

-Looking at the graphics above, we can tell that there are not many shots taken past half court. However, of the few shots that are taken past half court, a very low percentage of these shots are made. The percentage of shots made past half court for the 2000 and 2020 seasons is 0%. For both of those seasons, no shots taken past half court were made. Looking at the 2010 season, we can see that 1 of nines shots that were attempted were made past half court. Therefore, for the 2010 season, the percentage of shots made past half court was 11%. This is also a low percentage, showing us that the percentage of shots made past half court is low. This tells us that it is difficult to make shots further away from the basket, which is shown by the data.

-Based on the graphic displayed above, we can see that general trend that many shots attempted close to the basket were made shots. Looking at all three seasons, we can see that many of the data points close to the basket are green, meaning they were made baskets. This shows us that make shots closer to the basketball hoop is easier. This makes sense because there is less room for error because it is closer to the basketball hoop. 

-Lastly, the graphics above can provide some insight as to how defense has changed throughout the seasons. For example, it is reasonable to assume that the defense played tighter to their players in the 2000 season because many shots were taken from inside the three point line. During the 2010 season, we begin to see that more three point shots are taken. This could provide some insight that some of the Bucks opponents played close defense or that some of them played defense further off. The way that the defense was played shows how the shots begin to move closer to the three point arc. Lastly, in the 2020 season, we can see that more teams began to play defense further off of the Milwaukee Bucks and perhaps many of the Bucks opponents during this season played a zone defense. Therefore, we can see the relationship between offense and defense and how a defense is played can affect where the team shoots more from. This demonstrates how the graphic above can not only help our understanding of the offense but also provide some insight into the changes in defense throughout the years.

## Where should the NBA add in a 4 point arc? 

If the NBA were to include a 4 point arc, what distance should they place this arc? Using the expected value of shooting efficiency from each distance, we can put together the distance that the 4 point arc should be from the basketball hoop. I started by finding the distance of each shot from the basketball hoop using the Pythagorean theorem, which allows us to get the distance of each shot in feet. Next, using the the distances I had found and using the data whether each shot was made, I found the shooting efficiency of the players. This means how well the team could shot from each distance. Next, I calculated the expected value for each data range by using the shooting efficiency for each type of shot. After this, I combined the expected value for the two point shots and three point shots and used the maximum expected value for each distance range. Lastly, I used the expected value to find the maximum distance range that would give the highest expected value. This distance that was given was 28.284 feet. What does this distance mean? This distance is the optimal range for the 4 point arc to be from the basketball hoop to the top of the 4 point arc. 


